Serum Biomarkers of Nutrition Pre and Post CFTR Modulator Use in Children with Cystic Fibrosis

### 2022-04-07 Edith Zemanick/CF Nutrition/cf_nutrition.Rmd
cf nutrition mcp models

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(readxl)
library(performance)
library(mcp)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
source("~/GitHub/CF/get_genotype_class_severity.R")
```

```{r data cleaning}
# Import
df = read.csv("~/Dropbox/Work/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv")
# Only those on effective modulators
on_effective = unique(df$Patient.ID[which(df[,'Modulator'] %in% c('Kalydeco','Trikafta'))])
df = df %>% filter(Patient.ID %in% on_effective)
# Non-highy efective to "None
df$Modulator[!df$Modulator %in% c('Kalydeco','Trikafta')] = "None"
# Before or after
df$Before_After = df$Age.at.Test..Years. < df$Age.at.Start..years.
df$Before_After = factor(df$Before_After,levels = c(FALSE,TRUE),labels = c("After","Before"))
# Count number of measures 
num_values = df %>% group_by(Patient.ID,Before_After) %>% 
  summarise(n = sum(!is.na(X25OH.Vitamin.D)),.groups = "drop") %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = n)
num_values$`NA` = NULL
colnames(num_values)[2:3] = c("vit_measures_after","vit_measures_before")
df = left_join(df,num_values,by = "Patient.ID")
# At least two annual nutritional measurements prior to first modulator start date
two_measures = num_values$Patient.ID[which(num_values$`vit_measures_before` >= 2)]
# At least one measurement of nutritional markers and one weight/height measurement >= three months post-modulator
time_after = df %>% 
  filter(Age.at.Test..Years. >= (Age.at.Start..years. + 0.25),!is.na(X25OH.Vitamin.D)) %>%
  pull(Patient.ID) %>% unique(.)
# Filter out those without above criteria, get first modulator
df = df %>% filter(Patient.ID %in% intersect(two_measures,time_after)) %>%
  group_by(Patient.ID) %>% 
  mutate(first_mod = head(Modulator[Modulator != "None"],1)) %>% ungroup()
# Genotype group
geno = read_excel("/Users/timvigers/Dropbox/Work/CF/CF Info/Sweat Tests from Internal Database_genotype class.xlsx")
geno = geno[,c('PatientID','genoRisk')]
df = left_join(df,geno,by = c('Patient.ID' = 'PatientID'))
# Time prior to first modulator
years = df %>% group_by(Patient.ID,Before_After) %>% 
  mutate(years_to_start_max = max(abs(Age.at.Test..Years. - Age.at.Start..years.))) %>% 
  select(Patient.ID,Before_After,years_to_start_max) %>% 
  ungroup() %>% distinct(.) %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = years_to_start_max)
years$`NA` = NULL
colnames(years)[2:3] = c('max_years_to_start','max_years_from_start')
df = left_join(df,years,by = "Patient.ID")
df$years = df$Age.at.Test..Years. - df$Age.at.Start..years.
```

@@

```{r results='asis'}
nutrition_vars = c("VitaminA_Retinol","X25OH.Vitamin.D","Vitamin.E.Alpha",
                   "BMI","FEV1...pred")
# Get the most recent 
paired_df = df %>% 
  select(Patient.ID,Before_After,years,first_mod,all_of(nutrition_vars)) %>% 
  filter(years >= 0.25 | years < 0) %>%
  mutate(years = round(years)) %>%
  group_by(Patient.ID,years) %>%
  fill(all_of(nutrition_vars),.direction = "downup") %>%
  drop_na(any_of(c("VitaminA_Retinol","FEV1...pred"))) %>%
  group_by(Patient.ID,Before_After) %>%
  slice_min(abs(years)) %>% distinct() %>% filter(row_number()==1) 
paired_df$Before_After = relevel(paired_df$Before_After,ref = "Before")

f = as.formula(paste0("Before_After ~ ",paste0(nutrition_vars,collapse = "+")))
p = paired(f,data = paired_df,id = Patient.ID,strata = first_mod)
summary(p,
        labelTranslations = list(VitaminA_Retinol = "Vitamin A Retinol",
                                 X25OH.Vitamin.D = "25OH-Vitamin D",
                                 Vitamin.E.Alpha = "FEV1 % Predicted",
                                 first_mod = "First Modulator"))
```

@@

```{r}
# Set up data again
df = read.csv("~/Dropbox/Work/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv")
df$years = df$Age.at.Test..Years. - df$Age.at.Start..years.
t = df %>% group_by(Patient.ID) %>% summarise(n = sum(!is.na(VitaminA_Retinol)))
```

```{r}
outcome = "VitaminA_Retinol"
piecewise_mod = function(outcome,data){
  data = data %>% select(Patient.ID,years,all_of(outcome)) %>% drop_na()
  plot = ggplot(data,aes_string(x = "years",y = outcome,group = "Patient.ID")) + 
    geom_point() + geom_line()
}
```

>>>>

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(readxl)
library(performance)
library(mcp)
library(lmerTest)
library(plotly)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
source("~/GitHub/CF/get_genotype_class_severity.R")
```

```{r data cleaning}
# Import
df = read.csv("~/Dropbox/Work/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv",na.strings = "")
# Only those on effective modulators
on_effective = unique(df$Patient.ID[which(df[,'Modulator'] %in% c('Kalydeco','Trikafta'))])
df = df %>% filter(Patient.ID %in% on_effective)
# Non-highy efective to "None
df$Modulator[!df$Modulator %in% c('Kalydeco','Trikafta')] = "None"
# Before or after
df$Before_After = df$Age.at.Test..Years. < df$Age.at.Start..years.
df$Before_After = factor(df$Before_After,levels = c(FALSE,TRUE),labels = c("After","Before"))
# Count number of measures 
num_values = df %>% group_by(Patient.ID,Before_After) %>% 
  summarise(n = sum(!is.na(X25OH.Vitamin.D)),.groups = "drop") %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = n)
num_values$`NA` = NULL
colnames(num_values)[2:3] = c("vit_measures_after","vit_measures_before")
df = left_join(df,num_values,by = "Patient.ID")
# At least two annual nutritional measurements prior to first modulator start date
two_measures = num_values$Patient.ID[which(num_values$`vit_measures_before` >= 2)]
# At least one measurement of nutritional markers and one weight/height measurement >= three months post-modulator
time_after = df %>% 
  filter(Age.at.Test..Years. >= (Age.at.Start..years. + 0.25),!is.na(X25OH.Vitamin.D)) %>%
  pull(Patient.ID) %>% unique(.)
# Filter out those without above criteria, get first modulator
df = df %>% filter(Patient.ID %in% intersect(two_measures,time_after)) %>%
  group_by(Patient.ID) %>% 
  mutate(first_mod = head(Modulator[Modulator != "None"],1)) %>% ungroup()
# Genotype group
geno = read_excel("/Users/timvigers/Dropbox/Work/CF/CF Info/Sweat Tests from Internal Database_genotype class.xlsx")
geno = geno[,c('PatientID','genoRisk')]
df = left_join(df,geno,by = c('Patient.ID' = 'PatientID'))
# Time prior to first modulator
years = df %>% group_by(Patient.ID,Before_After) %>% 
  mutate(years_to_start_max = max(abs(Age.at.Test..Years. - Age.at.Start..years.))) %>% 
  select(Patient.ID,Before_After,years_to_start_max) %>% 
  ungroup() %>% distinct(.) %>%
  pivot_wider(id_cols = Patient.ID,names_from = Before_After,values_from = years_to_start_max)
years$`NA` = NULL
colnames(years)[2:3] = c('max_years_to_start','max_years_from_start')
df = left_join(df,years,by = "Patient.ID")
df$years = df$Age.at.Test..Years. - df$Age.at.Start..years.
```

@@

```{r results='asis'}
nutrition_vars = c("VitaminA_Retinol","X25OH.Vitamin.D","Vitamin.E.Alpha",
                   "BMI","FEV1...pred")
# Get the most recent 
paired_df = df %>% 
  select(Patient.ID,Before_After,years,first_mod,all_of(nutrition_vars)) %>% 
  filter(years >= 0.25 | years < 0) %>%
  mutate(years = round(years,1)) %>%
  group_by(Patient.ID,years) %>%
  fill(all_of(nutrition_vars),.direction = "downup") %>%
  drop_na(any_of(c("X25OH.Vitamin.D"))) %>%
  group_by(Patient.ID,Before_After) %>%
  slice_min(abs(years)) %>% distinct() %>% 
  arrange(Patient.ID,abs(years)) %>%filter(row_number()==1) 
paired_df$Before_After = relevel(paired_df$Before_After,ref = "Before")

f = as.formula(paste0("Before_After ~ ",paste0(nutrition_vars,collapse = "+")))
p = paired(f,data = paired_df,id = Patient.ID)
summary(p,
        labelTranslations = list(VitaminA_Retinol = "Vitamin A Retinol",
                                 X25OH.Vitamin.D = "25OH-Vitamin D",
                                 Vitamin.E.Alpha = "FEV1 % Predicted",
                                 first_mod = "First Modulator"))
```

@@

```{r}
# Set up data again
df = read.csv("~/Dropbox/Work/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_dataset.csv")
df$Age.at.Start..years. = NULL
# New age at start
mods = read_excel("~/Dropbox/Work/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
mods = mods[mods$Modulator %in% c('Kalydeco','Trikafta'),]
colnames(mods) = make.names(colnames(mods))
mods = mods %>% select(Patient.ID,Modulator,Age.at.Start..years.) %>% 
  pivot_wider(id_cols = Patient.ID,names_from = Modulator,
              values_from = Age.at.Start..years.)
colnames(mods)[2:ncol(mods)] = paste0("age_at_start_",colnames(mods)[2:ncol(mods)])
# Age at first and second modulator
mods$age_first = pmin(mods$age_at_start_Trikafta,mods$age_at_start_Kalydeco,na.rm = T)
mods$age_second = pmax(mods$age_at_start_Trikafta,mods$age_at_start_Kalydeco)
# Merge back in
df = left_join(df,mods,by = "Patient.ID")
df = df %>% group_by(Patient.ID) %>% 
  fill(age_at_start_Trikafta,age_at_start_Kalydeco,.direction = "downup")
# Time relative to first mod
df$time = df$Age.at.Test..Years. - df$age_first
# Calculate spline things
df$c1 = 
  (df$time-0)*(df$time>=0)
df$c2 = 
  (df$Age.at.Test..Years.-df$age_second)*(df$Age.at.Test..Years.>=df$age_second)
```

```{r}
piecewise_mod = function(outcome,data){
  data = data %>% select(Patient.ID,all_of(outcome),time,c1,c2) %>% drop_na()
  # Base plot
  plot = ggplot(data,aes_string(x = "time",y = outcome)) + 
    geom_point(alpha = 0.2) + geom_line(alpha = 0.2) +
    theme_bw() + theme(legend.position = "none") + xlab("Age at Test")
  # Model
  f = as.formula(paste0(outcome,"~ time + c1 + c2 + (1|Patient.ID)"))
  m = lmer(f,data = data)
  # Predicted values for plotting
  pred = predict(m)

  plot + geom_line(data=data, 
            aes(y = pred, x=Age.at.Test..Years.), size = 1, col="blue") 
  ggplotly(plot)
}
```

## Vitamin A

```{r}
mod_df = df %>% select(Patient.ID,VitaminA_Retinol,Age.at.Test..Years.,
                       age_first,age_second) %>% drop_na()
mod_df$c1 = 
  (mod_df$Age.at.Test..Years.-mod_df$age_first)*(mod_df$Age.at.Test..Years.>=mod_df$age_first)
mod_df$c2 = 
  (mod_df$Age.at.Test..Years.-mod_df$age_second)*(mod_df$Age.at.Test..Years.>=mod_df$age_second)

m = lmer(VitaminA_Retinol ~ c1+c2+(1|Patient.ID),data = mod_df)
```

```{r cache=TRUE}
mod_df = df %>% select(Patient.ID,time,VitaminA_Retinol) %>% drop_na()
model = list(
  VitaminA_Retinol ~ 1,  # Plateau in the first segment (int_1)
  1 + (1|Patient.ID) ~ 0 + time,    # Joined slope (time_2) at cp_1
  1 + (1|Patient.ID) ~ 0 + time     # Joined slope (int_3, time_3) at cp_2
)
fit = mcp(model, data = mod_df)
plot(fit) + theme_bw()
```

```{r eval=FALSE}
library(MultiKink)
data("triceps")
pred7 <- lm(triceps~ age + I((age-5)*(age>=5)) +
                                   I((age-10)*(age >= 10)),
                    data = triceps)
triceps$c1 = (triceps$age-5)*(triceps$age>=5)
triceps$c2 = (triceps$age-10)*(triceps$age>=10)

tri.age.plot <- ggplot(triceps, aes(x=age, y=triceps)) +
                 geom_point(alpha=0.55, color="black") + 
                 theme_minimal()
tri.age.plot +
  geom_line(data=triceps, 
            aes(y = pred7, x=age), size = 1, col="blue") 
```
